\boldmath PyUltraLight: A Pseudo-Spectral Code for the Dynamics of Ultra-Light Dark Matter

### 2018-05-09 draft.tex
Initial ShareLaTeX Import

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{float}

@@

Following \cite{Schwabe2016}, we consider the head-on collision of two solitions with mass ratio $\mu=2$ and high relative velocity. Using equation \ref{eq:covariant-solution}, we can write the total wavefunction of the binary system in terms of dimensionless quantities along the collision axis as
\begin{align}
    \psi(x,t)=& \ \alpha_1 f\big(\sqrt{\alpha_1}\vert x-x_1-v_1 t\vert\big)e^{i\left(\alpha_1\beta t+v_1(x-x_1)-\frac{1}{2}v_1^2 t+\delta\right)}\nonumber\\
    &+\alpha_2 f\big(\sqrt{\alpha_2}\vert x-x_2-v_2 t\vert\big)e^{i\left(\alpha_2\beta t+v_2(x-x_2)-\frac{1}{2}v_2^2 t\right)}
\end{align}
where $x_1$ and $x_2$ are the initial positions of the centres of the solitions, $v_1$ and $v_2$ are the velocities of the solitons, $\delta$ is a constant relative phase term, and $\alpha_1=1/4 \ \alpha_2$ parameterise the density profiles as discussed in section \ref{sec:soliton-profiles}. Because of the symmetry of the configuration, we take $v_1=-v_2$ and $x_1=-x_2$. We are interested in the density profile of the system at the time of maximal interference, where we expect the superposed wavefunction to possess pronounced features. Therefore, we choose the time at which the two components of the wavefunction are centred at the same location, such that $x_1+v_1t=-x_2-v_2t=0$. This corresponds to a time $t_{max}=\vert x_1/v_1\vert = \vert x_2/v_2\vert$. The dimensionless density is then given by
\begin{align}\label{eq:predicted-interference}
    \vert\psi(x,t_{max})\vert^2= \ \alpha_1^2\bigg[&f(\sqrt{\alpha_1}x)^2+16f(2\sqrt{\alpha_1}x)^2+\nonumber\\
    &8f(\sqrt{\alpha_1}x)f(2\sqrt{\alpha_1}x)\operatorname{cos}\left(-3\alpha_1\beta\left\vert\frac{x_1}{v_1}\right\vert+2v_1x+\delta \right)\bigg]
\end{align}
In Figure \ref{fig:interference}, we plot the dimensionless density profile at the time of maximal interference for two solitons with mass ratio $\mu=2$, and phase difference $\delta=0$. We see that the numerical result obtained using P\textsc{y}U\textsc{ltra}L\textsc{ight} closely matches the theoretical prediction of equation \ref{eq:predicted-interference}. Small disparities between the numerical and theoretical profiles may be attributed to the effect of gravitational contraction not included in the theoretical prediction of equation \ref{eq:predicted-interference}, and to a small offset in the true time of maximal interference. Neglecting these small deviations, we can verify that the code is well suited to faithfully reproduce the wave interference effects of the ULDM model.

\begin{figure}[tbp]
\centering % \begin{center}/\end{center} takes some additional vertical space
\includegraphics{interference_patterns}
\label{fig:interference}
\caption{Comparison of theoretical and numerical density profiles at time of maximal interference for head-on collision of two solitons with mass ratio $\mu=2$ and no relative phase difference. The solitons were chosen with dimensionless masses 5 and 10, with an initial separation of 4 code units and relative velocity of 20 code units. The simulation was run at $256^3$ for a box of side-length 8 code units.}
\end{figure}

@@

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM can give rise to effective forces which can dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, and do not require the ULDM model to incorporate explicit local interactions. Figure \ref{fig:repulsion} shows the results of two simulations using P\textsc{y}U\textsc{ltra}L\textsc{ight} in which two solitons of equal mass undergo a head-on collision. Contours of the density profile along the plane of symmetry are displayed. In the first simulation, there is no phase offset between the initial solitons, while in the second simulation the phases differ by $\pi$. We can see that the result of this phase shift is to create an effective repulsive force between the two solitons as they approach one another. Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.

\begin{figure}[tbp]
\centering % \begin{center}/\end{center} takes some additional vertical space
\includegraphics[width=1.\textwidth,trim=0 0 0 0,clip]{phase_comparison}
\caption{Head-on collisions between two solitons of mass 5, relative velocity 6, and initial separation 4 in code units. Time progresses from left to right across each row.}\label{fig:repulsion}
\end{figure}

% \subsection{Soliton Mergers}

% Want to show radial profile looking like NFW at the outer edges. See http://zulassung.hlrn.de/kurzbeschreibungen/nip00040.pdf

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018}. Here we reproduce just one of their results as a validity check for P\textsc{y}U\textsc{ltra}L\textsc{ight}. In particular, it is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann -S ellipsoid when orbiting a host mass. Figure \ref{fig:riemann} shows the internal velocity field of a solitonic satellite after one complete revolution around a host mass, obtained using P\textsc{y}U\textsc{ltra}L\textsc{ight} on a $256^3$ simulation grid. In this case the ratio of host to satellite mass was $\approx 55$. It can be seen that the soliton has become elongated along the radial line connecting to the host, indicating that it is tidally locked. Meanwhile the internal velocity field can be seen to be that of an irrotational Riemann-S ellipsoid \cite{Daller2012}.

\begin{figure}[tbp]
\centering 
\includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
\caption{Solitonic core after one revolution of a central potential, demonstrating the internal velocity field of an irrotational Riemann-S ellipsoid (where the collective motion of the soliton has been subtracted from the velocity field). Deformation of the spherical density profile into that of an ellipsoid can be seen, with extension along the radial direction (red arrow).}\label{fig:riemann}
\end{figure}

@@

\begin{figure}[tbp]
\centering 
\includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
\caption{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.}\label{fig:combined_1}
\end{figure}

\begin{figure}[tbp]
\centering 
\includegraphics[width=1.\textwidth,trim=0 0 2cm 2cm,clip]{energy_diagram2}
\caption{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.}\label{fig:binary}
\end{figure}

>>>>

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}

\usepackage[justification=justified]{caption}
\usepackage{varwidth}
\usepackage{graphicx}

@@

Following \cite{Schwabe2016}, we consider the head-on collision of two solitions with mass ratio $\mu=2$ and high relative velocity. Using equation \ref{eq:covariant-solution}, we can write the total wavefunction of the binary system in terms of dimensionless quantities along the collision axis as
\begin{align}
    \psi(x,t)=& \ \alpha_1 f\big(\sqrt{\alpha_1}\vert x-x_1-v_1 t\vert\big)e^{i\left(\alpha_1\beta t+v_1(x-x_1)-\frac{1}{2}v_1^2 t+\delta\right)}\nonumber\\
    &+\alpha_2 f\big(\sqrt{\alpha_2}\vert x-x_2-v_2 t\vert\big)e^{i\left(\alpha_2\beta t+v_2(x-x_2)-\frac{1}{2}v_2^2 t\right)}
\end{align}
where $x_1$ and $x_2$ are the initial positions of the centres of the solitions, $v_1$ and $v_2$ are the velocities of the solitons, $\delta$ is a constant relative phase term, and $\alpha_1=1/4 \ \alpha_2$ parameterise the density profiles as discussed in section \ref{sec:soliton-profiles}. Because of the symmetry of the configuration, we take $v_1=-v_2$ and $x_1=-x_2$. We are interested in the density profile of the system at the time of maximal interference, where we expect the superposed wavefunction to possess pronounced features. Therefore, we choose the time at which the two components of the wavefunction are centred at the same location, such that $x_1+v_1t=-x_2-v_2t=0$. This corresponds to a time $t_{max}=\vert x_1/v_1\vert = \vert x_2/v_2\vert$. The dimensionless density is then given by
\begin{align}\label{eq:predicted-interference}
    \vert\psi(x,t_{max})\vert^2= \ \alpha_1^2\bigg[&f(\sqrt{\alpha_1}x)^2+16f(2\sqrt{\alpha_1}x)^2+\nonumber\\
    &8f(\sqrt{\alpha_1}x)f(2\sqrt{\alpha_1}x)\operatorname{cos}\left(-3\alpha_1\beta\left\vert\frac{x_1}{v_1}\right\vert+2v_1x+\delta \right)\bigg]
\end{align}
In Figure \ref{fig:interference}, we plot the dimensionless density profile at the time of maximal interference for two solitons with mass ratio $\mu=2$, and phase difference $\delta=0$. We see that the numerical result obtained using P\textsc{y}U\textsc{ltra}L\textsc{ight} closely matches the theoretical prediction of equation \ref{eq:predicted-interference}. Small disparities between the numerical and theoretical profiles may be attributed to the effect of gravitational contraction not included in the theoretical prediction of equation \ref{eq:predicted-interference}, and to a small offset in the true time of maximal interference. Neglecting these small deviations, we can verify that the code is well suited to faithfully reproduce the wave interference effects of the ULDM model. 
\begin{center}
  \includegraphics[trim={0 0 0 0.9cm},clip, scale=0.9]{interference_patterns}
  \captionof{figure}{Comparison of theoretical and numerical density profiles at time of maximal interference for head-on collision of two solitons with mass ratio $\mu=2$ and no relative phase difference. The solitons were chosen with dimensionless masses 5 and 10, with an initial separation of 4 code units and relative velocity of 20 code units. The simulation was run at $256^3$ for a box of side-length 8 code units.}
  \label{fig:interference}
\end{center}

@@

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM can give rise to effective forces which can dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, and do not require the ULDM model to incorporate explicit local interactions. Figure \ref{fig:repulsion} shows the results of two simulations using P\textsc{y}U\textsc{ltra}L\textsc{ight} in which two solitons of equal mass undergo a head-on collision. Contours of the density profile along the plane of symmetry are displayed. In the first simulation, there is no phase offset between the initial solitons, while in the second simulation the phases differ by $\pi$. We can see that the result of this phase shift is to create an effective repulsive force between the two solitons as they approach one another. Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.
\begin{center}
  \includegraphics[width=1.\textwidth, trim={0 0 0 0},clip]{phase_comparison}
  \captionof{figure}{Head-on collisions between two solitons of mass 5, relative velocity 6, and initial separation 4 in code units. Time progresses from left to right across each row.}
  \label{fig:repulsion}
\end{center}

\vspace{1em}

@@

An extensive study of the tidal disruption of ULDM solitonic cores orbiting a central potential has recently been undertaken in \cite{Du2018}. Here we reproduce just one of their results as a validity check for P\textsc{y}U\textsc{ltra}L\textsc{ight}. In particular, it is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann -S ellipsoid when orbiting a host mass. Figure \ref{fig:riemann} shows the internal velocity field of a solitonic satellite after one complete revolution around a host mass, obtained using P\textsc{y}U\textsc{ltra}L\textsc{ight} on a $256^3$ simulation grid. In this case the ratio of host to satellite mass was $\approx 55$. It can be seen that the soliton has become elongated along the radial line connecting to the host, indicating that it is tidally locked. Meanwhile the internal velocity field can be seen to be that of an irrotational Riemann-S ellipsoid \cite{Daller2012}.
\begin{center}
  \includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
  \captionof{figure}{Solitonic core after one revolution of a central potential, demonstrating the internal velocity field of an irrotational Riemann-S ellipsoid (where the collective motion of the soliton has been subtracted from the velocity field). Deformation of the spherical density profile into that of an ellipsoid can be seen, with extension along the radial direction (red arrow).}
  \label{fig:riemann}
\end{center}

\vspace{1em}

@@

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
% \caption{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.}\label{fig:combined_1}
% \end{figure}

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.\textwidth,trim=0 0 2cm 2cm,clip]{energy_diagram2}
% \caption{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.}\label{fig:binary}
% \end{figure}

\begin{center}
  \includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
  \captionof{figure}{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.}
  \label{fig:combined_1}
\end{center}

\begin{center}
  \includegraphics[width=0.9\textwidth,trim=0 0 2cm 2cm,clip]{energy_diagram2}
  \captionof{figure}{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.}
  \label{fig:binary}
\end{center}
